---
title: "Inference for numerical data"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data, eval=TRUE}
load("more/nc.RData")
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

1.  What are the cases in this data set? How many cases are there in our sample?
--vb--
Cases are birth of child.
there are `r nrow(nc)` in this data set.

As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` command:

```{r summary, eval=FALSE}
summary(nc)
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

---vb variables:
  
variable      | type  
--------------|---------------------
fage          | numerical  
mage          | numerical  
mature        | categorical  
weeks         | numerical  
premie        | categorical  
visits        | numerical, discreet
marital       | categorical  
gained        | numerical  
weight        | numerical  
lowbirthweight| categorical  
gender        | categorical  
habit         | categorical  
whitemon      | categorical  

From the summary information, it appears that weeks and weight may have some outliers.  
To confirm, we will draw histogram for following variables: fage, weeks, and  weight.

```{r histograms, eval=TRUE}
library(ggplot2)
ggplot(nc, aes(x=mage)) + geom_histogram(binwidth = 4, fill="white", colour = "black")
ggplot(nc, aes(x=weeks)) + geom_histogram(binwidth = 2, fill="white", colour = "black")
ggplot(nc, aes(x=weight)) + geom_histogram(binwidth = 1, fill="white", colour = "black")
```
From the histograms, it appears that the distribution for variable weight and weeks are skewed to the left.   
They may be some outliers with very low and also very weights (above 10 less than 2). The distribution for "mage" is fairly normal.  

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?

--vb--
From Summary function, we could see that we have 1 case where value for habit was not specified.  We are ommiting this case.

```{r boxplots, eval=TRUE}
ggplot(nc[complete.cases(nc), ], aes(x=habit, y=weight)) + geom_boxplot()
      
```
  
From the plots, we can tell that the mean for the population of non-smokers has a value of median higher than for the smokers population.  Also, they are more possible ouliers for the non-smokers population.  Finally, the IQR for non-smokers is slightly less than for smokers.  

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following function to 
split the `weight` variable into the `habit` groups, then take the mean of each
using the `mean` function.

```{r by-means, eval=FALSE}
by(nc$weight, nc$habit, mean)
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test
.

## Inference

3.  Check if the conditions necessary for inference are satisfied. Note that 
you will need to obtain sample sizes to check the conditions. You can compute 
the group size using the same `by` command above but replacing `mean` with 
`length`.  

--vb--
```{r, sample_size, eval=TRUE}
by(nc$weight, nc$habit, length)
```

* The sample was randomly selected  
* we can assume that the sample (1000 cases) represent less than 10% of population we can assume independence of cases    
* Although there is a skewedness for the variable weight, the sample size is quite large (over 100 for each set somkers / non-smokers)  

4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.

--vb--
H0 = There is no difference in the average birth weight in new borns from mothers who did not smoke and from mothers who did. (${ \mu  }_{ ns }\quad -\quad { \mu  }_{ s }\quad =\quad 0$)
Ha = There is a difference in the average birth weight in new borns from mothers who smoke and from mothers who did not smoke. (${ \mu  }_{ ns }\quad -\quad { \mu  }_{ s }\quad \neq \quad 0$)  

Next, we introduce a new function, `inference`, that we will use for conducting
hypothesis tests and constructing confidence intervals. 

```{r inf-weight-habit-ht, eval=FALSE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

Let's pause for a moment to go through the arguments of this custom function. 
The first argument is `y`, which is the response variable that we are 
interested in: `nc$weight`. The second argument is the explanatory variable, 
`x`, which is the variable that splits the data into two groups, smokers and 
non-smokers: `nc$habit`. The third argument, `est`, is the parameter we're 
interested in: `"mean"` (other options are `"median"`, or `"proportion"`.) Next
we decide on the `type` of inference we want: a hypothesis test (`"ht"`) or a 
confidence interval (`"ci"`). When performing a hypothesis test, we also need 
to supply the `null` value, which in this case is `0`, since the null 
hypothesis sets the two population means equal to each other. The `alternative`
hypothesis can be `"less"`, `"greater"`, or `"twosided"`. Lastly, the `method` 
of inference can be `"theoretical"` or `"simulation"` based.

5.  Change the `type` argument to `"ci"` to construct and record a confidence 
interval for the difference between the weights of babies born to smoking and 
non-smoking mothers.

```{r interval, eval=TRUE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical")
```


By default the function reports an interval for ($\mu_{nonsmoker} - \mu_{smoker}$)
. We can easily change this order by using the `order` argument:

```{r inf-weight-habit-ci, eval=FALSE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical", 
          order = c("smoker","nonsmoker"))
```

* * *

## On your own

when looking at the summary statistics, there are NA's in some of the variables of interest
for rest of analysis.  First we will take a look at all the rows with at least one na.

```{r, eval = TRUE, echo = TRUE}
nc_na <- nc[!complete.cases(nc), ]
dim(nc_na)
```
By the results, it is clear that if we indescriminately remove all "incomplete" rows we would remove 200 row (out of 1000).  
This is too large a size and may impact some of the results. The variable "fage"" is the one with the most na's (171). 
Since we are not using this variable in the rest of the analysis, we will drop it from our data set.  Once this is done, 
we will filter out any row with a value of na in any of the remaining variables (38 out of 1000).  We will perform the analysis
this new data set.

```{r, eval = TRUE, echo = TRUE}
nc_fage_drop <- nc[, 2:13]
str(nc_fage_drop)

# Check for na on new data frame
nc_clean <- nc_fage_drop[complete.cases(nc_fage_drop), ]
dim(nc_clean)
```

-   Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. Note that since you're doing inference 
on a single population parameter, there is no explanatory variable, so you can 
omit the `x` variable from the function.


```{r, confidence_level_95, eval=TRUE}
inference(y=nc_clean$weeks, est="mean", type = "ci", conflevel=0.95, method="theoretical")
```
We are 95% confident that the average length of pregnancy (in weeks) is between 38.2162 and   
38.5738.  

-   Calculate a new confidence interval for the same parameter at the 90% 
confidence level. You can change the confidence level by adding a new argument 
to the function: `conflevel = 0.90`.
```{r, confidence_level_90, eval = TRUE}
inference(y=nc_clean$weeks, est="mean", type = "ci", conflevel=0.90, method="theoretical")
```

-   Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

H0: There is no difference between the average weight gain of young mothers and 
the average weight gain of mature mothers.  
Ha: There is a difference between the average weight gain of young mothers and 
the average weight gain of mature mothers.  

```{r, weight_gain, eval = TRUE}

ggplot(nc_clean, aes(x=mature, y=gained)) + geom_boxplot()

by(nc_clean$gained, nc_clean$mature, mean)

inference(y = nc_clean$gained, x = nc_clean$mature, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

-   Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.
```{r, eval=TRUE}
library(dplyr)
nc_clean %>% group_by(mature) %>% 
       summarise(min(mage), max(mage))
```
Age cut-off for mature mon = 35
Age cut-off for younger mon = 34

The function above will first group "mage"" by variable "mature"" then within each group, the minimum and maximum will be found.  
we are expecting that there is no overlap between the 2.  Hence, maximum of younger mothers and minimum of mature mothers are the cut-off...  

-   Pick a pair of numerical and categorical variables and come up with a 
research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using the `inference` 
function, report the statistical results, and also provide an explanation in 
plain language.

Does average number of hospital visits during pregnancy varies whether the mother is white or non-white?

First we will look at a 95% confidence interval for the variable "visits":
```{r, eval = TRUE, echo = TRUE}
inference(y=nc_clean$visit, est="mean", type = "ci", conflevel=0.95, method="theoretical")

```

H0: There is no difference between the average number of hospital visits for white mothers and non-white mothers
Ha: There is a difference between the average number of hostpital visits for white mothers and non-white mothers

```{r, visits, eval = TRUE}

ggplot(nc_clean, aes(x=whitemom, y=visits)) + geom_boxplot(fill = "khaki1")

by(nc_clean$visits, nc_clean$whitemom, mean)

inference(y = nc_clean$visits, x = nc_clean$whitemom, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

The p-value = 0.0312, is less that the significance level of $\alpha \quad =\quad 0.05$ and we reject the H0 hypothesis in favor of Ha hypothesis.

 

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was adapted for OpenIntro by Mine &Ccedil;etinkaya-Rundel from a lab 
written by the faculty and TAs of UCLA Statistics.
</div>
