---
title: "Multiple linear regression"
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=TRUE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.  
    
--vb--
This is an observational study.  No experiment were set-up.  The observations were gathered by end of class evaluations and questionair answered by 6 students pertaining on the beauty assessment of the professors.

Because this is not an experiment, we cannot conclude casual relationship.  We would need to rephrase the question.

Question: Does beauty impact a professor evaluation or is the difference due to sampling variation?  
-----  

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?
```{r score_distribution, eval = TRUE}
library(ggplot2)
ggplot(evals, aes(x=score)) + geom_histogram(binwidth = 0.5)
```
--vb--  
From the histogram, we can see that the distribution of 'score' is skewed to the left, with the majority of the observation with score of between 4 and 5.  This is as expected, most students will provide feeback with positive evaluation (4 or 5).  

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualization (scatterplot, side-by-side boxplots, or 
    mosaic plot).  
--vb--  
We will consider the following variables: `cls_perc_eval` and `cls_level`.  We will plot percentage of sutdent that complete evaluation dependent on the class level.  

```{r plot_eval_cls_level, eval = TRUE}
plot(evals$cls_perc_eval ~ evals$cls_level)
```

The percentage of student completing the evaluation has less variability for upper level class (since the box plot is smaller) than for lower.  There is no clear apparent difference for the median.  There is a marked difference for the lower whiskers.  For upper level class, the lower whicker ends higher with outliers.  There are no outiers for lower class level.

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg, eval = TRUE}
plot(evals$score ~ evals$bty_avg)
```


Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

```{r data_frame, eval = TRUE}
dim(evals)

summary(evals)
```

--vb--  
There are 463 rows in the data frame, each representing an observation.  Also, from the summary function results, there does not seem any missing values that would be represented as 'NA'.  We would expect the scatter plot to have 463 points.  However, it appears that the scatter plot only have 250 points represented.  
-----  

4.  Replot the scatterplot, but this time use the function `jitter()` on the
    $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was
    misleading about the initial scatterplot?  
    
--vb--  
We will had a small amount of noise the 'score' variable to break any possible "ties".  

```{r, scatter-score-bty_avg_jitter, eval = TRUE}
plot(jitter(evals$score) ~ evals$bty_avg)
```

It appears that quite a large number of points had the same values for (x,y), hence, they could not be differenciated on the scatter plot.  By adding a small amount of noise on the score variable (y), we can now differentiate the points.  

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the line to your plot
    using `abline(m_bty)`. Write out the equation for the linear model and
    interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?  
    
```{r linear_regression, eval = TRUE}
m_bty<- lm(score ~ bty_avg, data = evals)
plot(jitter(evals$score) ~ evals$bty_avg)
abline(m_bty)

summary(m_bty)
```

The equation of the regression line:  
$\hat { y } =\quad { b }_{ 0 }\quad +\quad { b }_{ 1 }\quad \times \quad x\quad =\quad 3.88034\quad +\quad 0.06664\quad \times \quad x$  

*Slope interpretation:*    
When the average beauty score of the professor goes up by 1, we would expect the rating evaluation to go up by .06664.  

*Statistical Significance:*  
Looking at the p value for the t score, we can say that the average beauty score is statiscally significant p(>|t|) = 5.08 x 10^-5^. 

We would not however find this to be a practically significant predictor.  Since this must be evaluated additionally of the professor evaluation.  In our day and age, "evaluating beauty" may cause some controversy and possibly not all respondent would reply.  we would imagine that it may difficult to have evaluation across gender line (i.e. male evaluating male, female evaluating female). Finally, While the debats is on whether there are universal metrics of beauty, are cultural and individual elements.  Therefore, it would be difficult to accurately value this variable.  

R^2^ is about 3.3%, hence, 3.3% of evaluations can be predicted accurately with the model.  

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).  
    
```{r residuals, eval=FALSE}
plot(m_bty$residuals ~ evals$bty_avg)
abline(h = 0, lty = 4)  # adds a horizontal dashed line at y = 0

#Historgream
hist(m_bty$residuals)

# normal probability plot of the residuals
qqnorm(m_bty$residuals)
qqline(m_bty$residuals)

```
 --vb--  
 *Conditions for the least squares line*  
 Linearity:  
Though it is difficult to tell with so many points, but it appear that the data show a slightly linear positive linearity.  

Nearly Normal residuals:  
From the Histogram, the residuals show a slightly left skewed distribution.  The normal probability plot of the residuals shows that the points do not follow the line for upper quadriles.

Constant Variability: 
From the residual plot, we can observe that there seems to have constant variability.  

Independent observations: 
We do not have much information on how the sample was taken.  We can assume indenpendence of the observations.  

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel, eval = TRUE}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels, eval = FALSE}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender, eval = FALSE}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.
--vb--  
```{r residuals-avg_gender, eval = FALSE}
# Normal Probability Plot
qqnorm(m_bty_gen$residuals)
qqline(m_bty_gen$residuals)

# residual plot against each predictor variable
plot(m_bty_gen$residuals ~ evals$bty_avg)
abline(h = 0, lty = 4)  # adds a horizontal dashed line at y = 0

plot(m_bty_gen$residuals ~ evals$gender)
abline(h = 0, lty = 4)  # adds a horizontal dashed line at y = 0

# 4 plots: Resiual vs Fitted, Normal Probability Plot, Scale-Location, Residual vs Leverage
plot(m_bty_gen)

#Historgream
hist(m_bty_gen$residuals)

# Checking linearlidity
plot(jitter(evals$score) ~ evals$bty_avg)
plot(evals$score ~ evals$gender)

```

From the histogram of residuals, we can see that the residuals distribution is slightly skewed to the left.  Looking at the Normal Probability Plot for residuals, the residuals do not follow the lines for upper quadriles.  Finally, Residuals vs Fitted, show that it appears to be constant variability for residuals.

The conditions are reasonable although, residuals are showing possible outliers and the Normal Probability plot shows breakdown, specially in the upper quadriles.  

We will assume indedence, we have no information on how the sample was taken or whether we can look at the collection order of the residuals to show independence of residuals.  

-----  

8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?  
    
--vb--
Adjusted R^2^ = 0.055, The parameter estimage for bty_avg = 0.07416


Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines, eval = FALSE}
multiLines(m_bty_gen)
```

9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?
    
    $\hat { score } =\quad 3.74734\quad +\quad 0.07416\quad \times \quad beauty\_ avg\quad +\quad 0.17239\quad \times \quad gender\_ male$  
    
    For gender = Male, we will evaluate the equation with gender_male = 1.  In case, of female gender, we will substitute a 0.  
    
    $\hat { score } =\quad 3.74734\quad +\quad 0.07416\quad \times \quad beauty\_ avg\quad +\quad 0.17239$  
    
    Male professor will have a evaluation score higher by 0.17239 all other things being equal.  
    
The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.  
    
```{r m_bty_rank, eval = TRUE}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

For variable with more than 2 levels, it appears to handle it considerering them 2 different variables.  
The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.
--vb-- 
Possibly, number of professors teaching sections in course in sample: single, multiple, since the evaluation are done within a class/section.  Whether the professor is teaching multiple sections should not have an impact a given class evaluation.  

Let's run the model...

```{r m_full, eval = TRUE, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.  
    
    The p value for this variable, is 0.77806 and is the highest in the model.  

13. Interpret the coefficient associated with the ethnicity variable.    
    All other things being equal, Evaluation for professor that not minority tends to be 0.1234929 higher.  

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?  
    
```{r m_full_1, eval = TRUE, tidy = FALSE}
m_full_1 <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full_1)

m_full$coefficients - m_full_1$coefficients

```
The coefficients and significance changed slightly. Since the values changed, the drop variable was slightly collinear with the other explanatory variables.  

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.
    
```{r m_full_best, eval = TRUE, tidy = FALSE}
m_full_best <- lm(score ~ ethnicity + gender + language + age + cls_perc_eval 
             +   cls_credits + bty_avg 
             +  pic_color, data = evals)
summary(m_full_best)
```

$\hat { score } =\quad 3.771922\quad +\quad 0.051069\cdot beauty\_ avg\quad -\quad 0.006272\cdot age\quad +\quad 0.004524\cdot Stud%completeEval\quad +\quad 0.159237\cdot notminority\quad +\quad 0.211402\cdot gender\_ male\quad -0.221755\cdot langnon-english\quad +\quad 0.038055\cdot upperlvl\quad +\quad 0.528038\cdot onecrd\quad -\quad 0.202161\cdot colorpic$

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.
```{r residual_best_fit, eval = TRUE}
# Normal Probability Plot
qqnorm(m_full_best$residuals)
qqline(m_full_best$residuals)

#Historgream
hist(m_full_best$residuals) 

# 4 plots: Resiual vs Fitted, Normal Probability Plot, Scale-Location, Residual vs Leverage
plot(m_full_best)

# Checking linearlidity
plot(jitter(evals$score) ~ evals$bty_avg)
plot(jitter(evals$score) ~ evals$gender)
plot(jitter(evals$score) ~ evals$ethnicity)
plot(jitter(evals$score) ~ evals$gender)
plot(jitter(evals$score) ~ evals$language)
plot(jitter(evals$score) ~ evals$age)
plot(jitter(evals$score) ~ evals$cls_perc_eval)
plot(jitter(evals$score) ~ evals$cls_credits)
plot(jitter(evals$score) ~ evals$pic_color)
```

The residuals of the model are nearly model, although we have teh same breakdown as previously for upper quartiles when looking at normal qq plot.  
The Variability of the residuals is nearly constant (Residuals vs Fitted)  
Each variability is linearly related (scatter plot for age, bty_avg, cls_perc_eval)  
we will assume indedendence, we have limited information on how original sample was selected and we cannot infere independency of residuals by looking at order of collection.  

17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?  
    
--vb-- 
I am not sure I understand the question  
I would think not, I do not see how would this impact the model.  


18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.  
    
Professor would be younger male teaching one credit class, he would not belong to a minority group, he would have received this degree from a universtity where english is the language.  The professor would have a black and white picture and who have been rated beautifull.    

19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?  
    
No, this was not conducted as an experiment but based on a sample in a given university.  I would also argue that this may be time-based.  As cultural value changes, these results may be different in other university or in a different time frame.  

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). This lab was written by 
Mine &Ccedil;etinkaya-Rundel and Andrew Bray.
</div>
